Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-09-12 Writing/NatureWater/2ndRoundRevision/replies/reply_to_reviewers_R2.tex

  \begin{itemize}
  \item (1.a) Does the addition of ``X'' to Equation (2) explain the changed irrigated area results in Figure 3?

  \textcolor{blue}{Yes. Please note that the results for the share of irrigated acre in this manuscript are also different from the previous version of our manuscript  due to the code updates that we have summarized at the top of this document. All the qualitative results and conclusions regarding the impacts of aquifer thickness are the same as the previous versions of our manuscript.}

  \item (1.b) What motivates the other changes made to Equation (2)?

  \textcolor{blue}{The other changes made to Equation 2 include the addition of state-year FE, in order to provide stronger controls than just including year FE. In addition, we choose not to allow the interactions of $WB$ and $AT$. This decision was made in the interests of developing a more parsimonious model that considers the impacts of $WB$ and $AT$ individually. Our results remain virtually identical though whether this the interaction between $WB$ and $AT$ is included or not. }

@@

    \textcolor{blue}{In the previous version of our average productivity analysis, we assess the impacts of four aquifer thickness values: 10, 50, 90, and 130 meters in the initial version of our manuscript, with equal spacing chosen to illustrate non-linearity in impacts of aquifer thickness. In our last round of revisions, we changed these values to instead be 10, 40, 70, and 100m. These values ensure that we have a range of values spanning different aquifer thickness categories. Please note that key qualitative message and findings is identical whichever set of values we use. As we mentioned earlier, we made this change to illustrate that significant non-linear changes in total productivity can happen within a narrower range of evenly spaced aquifer thickness compared to the previous sets of aquifer thickness values used in this figure. This change also provide results that are more realistically centered on the range of aquifer thickness values observed within our study area and period.}

\item New comment 4 – causality and fixed effects: The manuscript has been strengthened by the expansion of the section explaining the reasons for assuming declining aquifer thickness causally reduces the ability to use irrigation to maintain crop yields (lines 40-57). Please clarify the following:

  \begin{itemize}
  \item (4.a) Please clarify for readers if other major causal pathways might be relevant. For example: Are there relevant regulations that limit abstractions in counties where aquifers are too depleted? If so, the causality would be similar, but the implications somewhat different. If not, then stating this would be helpful so readers can rule out potentially strong alternative causal relationships.

  \textcolor{blue}{Regulatory pumping restrictions have limited impacts on water use decisions in our study area, and therefore are unlikely to impact the ability to use groundwater as a buffer against drought. In contrast, the impacts of aquifer depletion on farmers' ability to meet crop water needs through changes in well yields are linked directly to hydrologic and agronomic theory, and have demonstrated extensively through prior work cited in our manuscript. While we note that restrictions in pumping are imposed in some parts of the High Plains Aquifer, there are several reasons why we can confidently state these would not be a confounding source of causality in our analysis. Please see the list below. We added the following paragraphs as Appendix A in this round of manuscript.}

  \begin{enumerate}
      \item \textcolor{blue}{The majority of groundwater withdrawal regulations that do exist in the HPA limit pumping to a specified volume over a multiyear period, and farmers have also been allowed to bank and carryover significant volumes of water deficits from historic periods when allocation levels were higher (Rimsaite and Brozovic, 2023). This enables farmers to increase water withdrawals substantially in drought years when demand is higher, and prior research has shown that farmers' water use decisions are not constrained by such regulations even in extreme drought years such as 2012 (Foster et al., 2019).}
      \item \textcolor{blue}{Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas (Marston et al., 2022; Rimsaite and Brozoic, 2023), these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in NW Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the LEMA area covers only parts of two counties and hence represents a very small proportion of our overall study sample while the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Moreover, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions (Drysdale and Hendricks, 2018; Deines et al., 2021). These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.}
      \item \textcolor{blue}{Areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction due to laws that typically provide farmers with absolute rights to use groundwater under their land coupled with low levels of metering that constrain potential for implementing extensive and binding pumping restrictions (Wheeler et al., 2016; Closas and Molle, 2018).}
  \end{enumerate}

@@

  \item (4.b) Relatedly, the responses to my previous comments 1.a and 7.b indicated that county-level fixed effects were not included in any regressions (only state-level). The response to reviewer \#2 and the Methods section suggest that county-level fixed effects were used in Equation 1. Please clarify the discrepancy and make sure this is clear in the paper.

  \textcolor{blue}{In our initial submission, we noted in our methods that county fixed effects $\alpha_i$ were included in equation 1 but not in equation 2. The reason for this difference is outlined in Appendix C, which provides a detailed discussion of why including county fixed effects for our irrigated area share regression is problematic.}

  \item (4.c) The following conclusion in Appendix C, lines 410-414 does not seem fully convincing in its present form: The fact that state-level fixed effects are not influential does not immediately imply that county-level fixed effects are ``unlikely to cause significant bias''. There could e.g. easily be relevant county-level effects that average out at the state level.

  \textcolor{blue}{We recognize your point that this may be potentially too strong of a claim. To more carefully summarise the implications of our choice of fixed effects, we have modified the text on lines 430-435 for our revision of the statement.}

@@

\item New comment 5 – Figure 1:

  \begin{itemize}
   \item (5.a) The histograms are a helpful addition to Figure 1. Please consider adding the total number of observations to the caption to help readers interpret the values. I also recommend adding such histograms to Figures 3 and 4 for consistency. Please clarify why the histogram reaches further than the curves in Figure 1.B and correct this if necessary.

   \textcolor{blue}{We have added total number of observations to the captions of Figures 1 and 3. Figure 4 is the results of simulation based on the estimated model represented by equations 1 and 2, and therefore there is no appropriate number of observations for the figure so the value is omitted.}

   \textcolor{blue}{We have added histograms for Figures 3 and 4. The inconsistency in the range of the histograms and the main figures were caused by a small coding error that did not correctly align two figures vertically. These problems are now fixed. Please note that because of the width (range) of the histogram bins, it looks like the bins at the left and right ends are beyond the range of the main figure, but this is not the case and is simply a plotting feature.}

@@

\begin{itemize}

\item (M1) The response letter indicates that ``all [minor] points have been addressed''. Most minor comments were addressed correctly. The following minor comments from the previous round, however, have not yet been addressed:

\item (M1.a) P. 15: ``WB'' and ``ST'' [now ``AT''] should not be in italics, unless ``W'', ``B'', ``S'', and ``T'' are separate variables

\textcolor{blue}{Instead of making $WB$ and $AT$ un-italicized, we simply replaced them with $D$ and $A$, respectively for all of their instances.}

### 2023-09-12 Writing/NatureWater/2ndRoundRevision/manuscript.tex
edits

Despite widespread alarm about the risks that aquifer depletion poses to crop productivity and food security, empirical evidence is limited about how reductions in groundwater availability will alter farmers capacity to adapt successfully to drought and rainfall variability. Most empirical studies that have assessed impacts of drought on agricultural productivity have focused primarily on rainfed production areas due to the greater exposure of rainfed agriculture to drought-related shocks \citep{schlenker2009nonlinear, lobell2014greater, schlenker2010robust, zhou2020connections, borgomeo2020impact}. Existing studies also demonstrate the importance of existing irrigated areas or future irrigation intensification and expansion to mitigate negative impacts of drought \citep{kuwayama2019estimating, zipper2016drought, zhu2022untangling, zhu2022warming, lu2020mapping, davis2019sensitivity, li2018changes}. In contrast, most research assessing impacts of aquifer depletion on resilience of irrigated farmland to drought has relied on simulation modeling \citep{foster2015well, cotterman2018groundwater, kahil2015modeling, yoon2021coupled, rad2020mod}, with more limited work that attempts to empirically evaluate change in drought-related production losses and risks on irrigated lands as a function of changing aquifer storage \citep{jain2021groundwater, suter2021depletion}.

@@

We now focus on comparing yield response to water deficit among the three aquifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \ref{fig:yield-response}). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers ability to fully meet crop water requirements during periods of more severe or extended water deficits \citetext{\citealp[\citet{foster2014modeling}]{rad2020effects}; \citealp{hrozencik2017heterogeneous}}. In contrast, yield continues to increase gradually as water deficit increases beyond 400 mm for counties with the highest level of aquifer thickness.

@@

While aquifer thickness has a statistically significant impact on per-area irrigated crop yields in drought conditions (water deficit greater than 700 mm), the magnitude of the yield impact is small when compared to differences between irrigated and rainfed production for the same weather conditions. One explanation for this result is that farmers may choose to retain a smaller proportion of land in irrigated production, so that limited groundwater pumping capacity can be used to adequately buffer crops against drought on land that remains under irrigation. Reductions in irrigated areas as an aquifer is depleted have been demonstrated theoretically \citep{rad2020effects, foster2014modeling, hrozencik2017heterogeneous, deines2020transitions}, but there has been little empirical research on farmers irrigated area choices or how these are affected by combinations and magnitudes of weather risks and groundwater scarcity.

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.09 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading in 1.57 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

\caption{Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity.}\label{fig:tot-impact}
\end{figure}

@@

We demonstrate the important role of aquifer thickness in determining farmers ability to successfully buffer crops against drought risks in groundwater irrigated agricultural systems such as the High Plains of the United States. Previous research focused on impacts of drought on agricultural systems has largely focused either on risks for rainfed production \citep{schlenker2009nonlinear, lobell2014greater, schlenker2010robust, zhou2020connections, borgomeo2020impact} or for irrigated lands where water supply constraints are assumed to be minimal \citep{kuwayama2019estimating, zipper2016drought, zhu2022untangling, zhu2022warming, lu2020mapping, davis2019sensitivity, li2018changes, luan2021combined}. Our analysis adds new empirical insights to this literature by quantifying how reductions in aquifer thickness increase the vulnerability of irrigated agricultural production areas to drought events. Critically, our findings suggest that commonly adopted binary definitions of agricultural land areas into rainfed and irrigated categories are likely to underestimate exposure of crop production to drought risks, in particular in areas where water supplies for irrigation are insufficient to fully meet crop water needs in all years due to constraints imposed by hydrology, policies, and/or economics.

@@

Our findings also provide useful guidance for water managers and policymakers seeking to address unsustainable groundwater use in the US High Plains and globally. Consistent with prior theoretical modeling, we demonstrate that the relationship between aquifer thickness and climatic water deficits during the growing season is highly non-linear when accounting for farmers decisions about irrigated production areas. This indicates that sustainable groundwater conservation efforts \citep{macewan2017hydroecon, butler2020charting, elshall2020groundwater} should be targeted in space and time to avoid depletion exceeding critical thresholds which impair farmers' ability to effectively buffer production against drought. Analysis such as ours can provide valuable evidence about these critical thresholds and tipping points to guide proactive management of groundwater stocks, and, in doing so, limit economically damaging transitions from irrigated to rainfed production \citep{foster2017effects, deines2020transitions}. For example, several groundwater management agencies in the Texas portion of the HPA have in recent years set out targets to ensure 50\% of current aquifer storage and thickness remains in 50 years time (so called 50/50 rule) \citep{closas2018chronicle}. The effectiveness of such targets could likely be improved substantially by setting conservation targets based on empirical evidence about the level of aquifer thickness that is required to ensure adequate levels of drought protection, both now and with future climate change.

@@

Second, our analysis considered the effects of aquifer thickness on farmers decisions to switch from irrigated to rainfed production, but did not assess other potential responses to aquifer depletion such as changing crop types or irrigation management practices. Switching to more drought resistant crops or varieties may enable farmers to maintain irrigated production at lower levels of aquifer thickness, and has been shown to be an important response to increasing agricultural water scarcity and abstraction policies in other studies \citep{bhattarai2021impact, deines2019quantifying, manning2017producer}. Improvements to irrigation scheduling practices and technologies, such as have been implemented as part of the SD6-LEMA (Local Enhanced Management Area) in Kansas \citep{deines2019quantifying, glose2022quantifying} and in parts of Texas \citep{mrad2020peak} and which are proposed as part of California's Sustainable Groundwater Management Act (SGMA) \citep{berbel2019droughts, lubell2020sustainable}, could also help to improve the efficiency of irrigation and, hence, the ability to more effectively meet crop water demands with lower well capacities. Such interventions may reduce - but not eliminate - increases in drought risk caused by lower or declining aquifer thickness, and could also help to mitigate against potential yield losses caused by over-use of water in wetter years via raising awareness of efficiency saving potentials \citep{foster2019assessing} and the importance of conservation \citep{marston2022}. However, risks also exist that irrigation efficiency improvements will increase net extraction, ultimately causing a rebound effect that exacerbates future aquifer depletion and drought risks \citep{grafton2018paradox,perez2021agricultural}. Future research should explore potential strategies that farmers could adopt to lessen the combined impacts of drought and aquifer depletion, alongside efforts to curb unsustainable groundwater extraction and climate change.

Finally, we focus on the impacts of aquifer thickness on agricultural resilience to seasonal water deficits. However, where water deficits are distributed unevenly during the season, negative impacts of aquifer depletion may be even larger. This is because reductions in well yields caused by aquifer depletion will constrain farmers ability to meet peak crop water demands, increasing yield losses compared with if water deficits had been distributed evenly throughout the season \citep{ortiz2019unpacking}. Furthermore, irrigation also has broader benefits for crops such as reducing heat-related stress through cooling of land surface temperatures \citep{adegoke2003impact, bonfils2007empirical, lobell2008effect, zhu2022untangling}. Reductions in aquifer thickness therefore may have other negative impacts on irrigated crop productivity, if constraints to water supply due to aquifer depletion are significant enough to limit cooling benefits of irrigation during the growing season.

@@

To estimate the additional impact of aquifer thickness on the share of total acres that are irrigated in a county, a fractional logit model under the generalized additive model framework is used in which the impact of aquifer thickness, average water deficit, and their interactions are estimated without the prior requirement to assume any functional forms. This allows for flexible representation of the non-linear relationships between drought risks, aquifer conditions, and agricultural land use, including potential thresholds in farmers irrigated area choices under water supply constraints that have been noted in prior work \citep{foster2015well, foster2015analysis, hrozencik2017heterogeneous}. The model is given in Equation \eqref{eq:eqn2} below:

@@

@@

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022importance, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in NW Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the LEMA area covers only parts of two counties and hence represents a very small proportion of our overall study sample while the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Moreover, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

>>>>

Despite widespread alarm about the risks that aquifer depletion poses to crop productivity and food security, empirical evidence is limited about how reductions in groundwater availability will alter farmers' capacity to adapt successfully to drought and rainfall variability. Most empirical studies that have assessed impacts of drought on agricultural productivity have focused primarily on rainfed production areas due to the greater exposure of rainfed agriculture to drought-related shocks \citep{schlenker2009nonlinear, lobell2014greater, schlenker2010robust, zhou2020connections, borgomeo2020impact}. Existing studies also demonstrate the importance of existing irrigated areas or future irrigation intensification and expansion to mitigate negative impacts of drought \citep{kuwayama2019estimating, zipper2016drought, zhu2022untangling, zhu2022warming, lu2020mapping, davis2019sensitivity, li2018changes}. In contrast, most research assessing impacts of aquifer depletion on resilience of irrigated farmland to drought has relied on simulation modeling \citep{foster2015well, cotterman2018groundwater, kahil2015modeling, yoon2021coupled, rad2020mod}, with more limited work that attempts to empirically evaluate change in drought-related production losses and risks on irrigated lands as a function of changing aquifer storage \citep{jain2021groundwater, suter2021depletion}.

@@

We now focus on comparing yield response to water deficit among the three aquifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \ref{fig:yield-response}). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers' ability to fully meet crop water requirements during periods of more severe or extended water deficits \citetext{\citealp[\citet{foster2014modeling}]{rad2020effects}; \citealp{hrozencik2017heterogeneous}}. In contrast, yield continues to increase gradually as water deficit increases beyond 400 mm for counties with the highest level of aquifer thickness.

@@

While aquifer thickness has a statistically significant impact on per-area irrigated crop yields in drought conditions (water deficit greater than 700 mm), the magnitude of the yield impact is small when compared to differences between irrigated and rainfed production for the same weather conditions. One explanation for this result is that farmers may choose to retain a smaller proportion of land in irrigated production, so that limited groundwater pumping capacity can be used to adequately buffer crops against drought on land that remains under irrigation. Reductions in irrigated areas as an aquifer is depleted have been demonstrated theoretically \citep{rad2020effects, foster2014modeling, hrozencik2017heterogeneous, deines2020transitions}, but there has been little empirical research on farmers' irrigated area choices or how these are affected by combinations and magnitudes of weather risks and groundwater scarcity.

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.09 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading in 1.57 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

\caption{Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity.}\label{fig:tot-impact}
\end{figure}

@@

We demonstrate the important role of aquifer thickness in determining farmers' ability to successfully buffer crops against drought risks in groundwater irrigated agricultural systems such as the High Plains of the United States. Previous research focused on impacts of drought on agricultural systems has largely focused either on risks for rainfed production \citep{schlenker2009nonlinear, lobell2014greater, schlenker2010robust, zhou2020connections, borgomeo2020impact} or for irrigated lands where water supply constraints are assumed to be minimal \citep{kuwayama2019estimating, zipper2016drought, zhu2022untangling, zhu2022warming, lu2020mapping, davis2019sensitivity, li2018changes, luan2021combined}. Our analysis adds new empirical insights to this literature by quantifying how reductions in aquifer thickness increase the vulnerability of irrigated agricultural production areas to drought events. Critically, our findings suggest that commonly adopted binary definitions of agricultural land areas into rainfed and irrigated categories are likely to underestimate exposure of crop production to drought risks, in particular in areas where water supplies for irrigation are insufficient to fully meet crop water needs in all years due to constraints imposed by hydrology, policies, and/or economics.

@@

Our findings also provide useful guidance for water managers and policymakers seeking to address unsustainable groundwater use in the US High Plains and globally. Consistent with prior theoretical modeling, we demonstrate that the relationship between aquifer thickness and climatic water deficits during the growing season is highly non-linear when accounting for farmers' decisions about irrigated production areas. This indicates that sustainable groundwater conservation efforts \citep{macewan2017hydroecon, butler2020charting, elshall2020groundwater} should be targeted in space and time to avoid depletion exceeding critical thresholds which impair farmers' ability to effectively buffer production against drought. Analysis such as ours can provide valuable evidence about these critical thresholds and tipping points to guide proactive management of groundwater stocks, and, in doing so, limit economically damaging transitions from irrigated to rainfed production \citep{foster2017effects, deines2020transitions}. For example, several groundwater management agencies in the Texas portion of the HPA have in recent years set out targets to ensure 50\% of current aquifer storage and thickness remains in 50 years time (so called 50/50 rule) \citep{closas2018chronicle}. The effectiveness of such targets could likely be improved substantially by setting conservation targets based on empirical evidence about the level of aquifer thickness that is required to ensure adequate levels of drought protection, both now and with future climate change.

@@

Second, our analysis considered the effects of aquifer thickness on farmers' decisions to switch from irrigated to rainfed production, but did not assess other potential responses to aquifer depletion such as changing crop types or irrigation management practices. Switching to more drought resistant crops or varieties may enable farmers to maintain irrigated production at lower levels of aquifer thickness, and has been shown to be an important response to increasing agricultural water scarcity and abstraction policies in other studies \citep{bhattarai2021impact, deines2019quantifying, manning2017producer}. Improvements to irrigation scheduling practices and technologies, such as have been implemented as part of the SD6-LEMA (Local Enhanced Management Area) in Kansas \citep{deines2019quantifying, glose2022quantifying} and in parts of Texas \citep{mrad2020peak} and which are proposed as part of California's Sustainable Groundwater Management Act (SGMA) \citep{berbel2019droughts, lubell2020sustainable}, could also help to improve the efficiency of irrigation and, hence, the ability to more effectively meet crop water demands with lower well capacities. Such interventions may reduce - but not eliminate - increases in drought risk caused by lower or declining aquifer thickness, and could also help to mitigate against potential yield losses caused by over-use of water in wetter years via raising awareness of efficiency saving potentials \citep{foster2019assessing} and the importance of conservation \citep{marston2022}. However, risks also exist that irrigation efficiency improvements will increase net extraction, ultimately causing a rebound effect that exacerbates future aquifer depletion and drought risks \citep{grafton2018paradox,perez2021agricultural}. Future research should explore potential strategies that farmers could adopt to lessen the combined impacts of drought and aquifer depletion, alongside efforts to curb unsustainable groundwater extraction and climate change.

Finally, we focus on the impacts of aquifer thickness on agricultural resilience to seasonal water deficits. However, where water deficits are distributed unevenly during the season, negative impacts of aquifer depletion may be even larger. This is because reductions in well yields caused by aquifer depletion will constrain farmers' ability to meet peak crop water demands, increasing yield losses compared with if water deficits had been distributed evenly throughout the season \citep{ortiz2019unpacking}. Furthermore, irrigation also has broader benefits for crops such as reducing heat-related stress through cooling of land surface temperatures \citep{adegoke2003impact, bonfils2007empirical, lobell2008effect, zhu2022untangling}. Reductions in aquifer thickness therefore may have other negative impacts on irrigated crop productivity, if constraints to water supply due to aquifer depletion are significant enough to limit cooling benefits of irrigation during the growing season.

@@

To estimate the additional impact of aquifer thickness on the share of total acres that are irrigated in a county, a fractional logit model under the generalized additive model framework is used in which the impact of aquifer thickness, average water deficit, and their interactions are estimated without the prior requirement to assume any functional forms. This allows for flexible representation of the non-linear relationships between drought risks, aquifer conditions, and agricultural land use, including potential thresholds in farmers' irrigated area choices under water supply constraints that have been noted in prior work \citep{foster2015well, foster2015analysis, hrozencik2017heterogeneous}. The model is given in Equation \eqref{eq:eqn2} below:

@@

Some parts of the HPA region face regulatory pumping restrictions in response to declines in aquifer thickness. Resulting groundwater use reductions can in turn negatively affect crop yield. This causal pathway of aquifer thickness affecting crop yield is distinct from the pathways via well yield discussed in the \ref{main} section. However, it is unlikely that our regression analysis capture this effect.

@@

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in NW Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the LEMA area covers only parts of two counties and hence represents a very small proportion of our overall study sample while the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Moreover, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

### 2023-09-12 Writing/NatureWater/2ndRoundRevision/replies/cover_letter.tex
edits

\begin{itemize}
    \item Fix a coding error that was creating invalid observations for our regressions analysis of the share of irrigated acres. Consequently, our results changed slightly. Our main conclusions and qualitative results are robust and did not change. 
    \item We added histograms of aquifer thickness in Figure 3.
\end{itemize}

>>>>

\begin{itemize}
    \item Fix a coding error that was creating invalid observations for our regressions analysis of the share of irrigated acres. Consequently, our results changed slightly. But, our main conclusions and qualitative results are robust and did not change. 
    \item We added histograms of aquifer thickness in Figure 3 as suggested by reviewer 3.
    \item We added another section in Appendix (Appendix A) that discusses groundwater regulations in response to declines in aquifer thickness as a causal pathway through which aquifer thickness affects yield. This is in response to a comment from reviewer 3.
\end{itemize}
